Researchers collaborate with same-gendered colleagues more often than expected across the life sciences

### 2018-10-29 manuscript/Manuscript_Claire.tex
Resubmission to PLOS ONE

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Researchers preferentially collaborate with same-gendered colleagues across the life sciences},
            pdfauthor={Luke Holman* and Claire Morandin; *;},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\setlength{\droptitle}{-2em}
  \title{Researchers preferentially collaborate with same-gendered colleagues
across the life sciences}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{Luke Holman* and Claire Morandin\textsuperscript{§} \\ *\textit{luke.holman@unimelb.edu.au} \\ \textsuperscript{§}\textit{claire.morandin@helsinki.fi} \vspace{5mm}}
  \preauthor{\centering\large\emph}
  \postauthor{\par}
  \date{}
  \predate{}\postdate{}

@@

\begin{document}
\maketitle
\begin{abstract}
Evidence suggests that women in academia are hindered by conscious and
unconscious biases, and often feel excluded from formal and informal
opportunities for research collaboration. In addition to ensuring
fairness and helping to redress gender imbalance in the academic
workforce, increasing women's access to collaboration could help
scientific progress by drawing on more of the available human capital.
Here, we test whether researchers preferentially collaborate with
same-gendered colleagues, using more stringent methods and a larger
dataset than in past work. Our results reaffirm that researchers
preferentially co-publish with colleagues of the same gender, and show
that this `gender homophily' is slightly stronger today than it was 10
years ago. Contrary to our expectations, we found no evidence that
homophily is driven mostly by senior academics, and no evidence that
homophily is stronger in fields where women are in the minority.
Interestingly, journals with a high impact factor for their discipline
tended to have comparatively low homophily, as predicted if mixed-gender
teams produce better research. We discuss some potential causes of
gender homophily in science. \vspace{5mm}
\par\noindent \textbf{Keywords:} Gender bias, Homophily, Research
collaboration, Text mining, Women in STEM.
\end{abstract}

\maketitle

\unmarkedfntext{*School of BioSciences, The University of Melbourne, Victoria, Australia.}
\maketitle\unmarkedfntext{\textsuperscript{§}Organismal and Evolutionary Biology Research Programme,
Faculty of Biological and Environmental Sciences, University of Helsinki, Finland.}

@@

\section{Introduction}\label{introduction}

@@

However, we believe that prior studies of gender homophily were hindered
by a largely unacknowledged statistical issue that we name the Wahlund
effect (\autoref{wahlund_plot}), by analogy with the conceptually
similar Whalund effect in population genetics {[}59{]}. The Wahlund
effect makes it deceptively difficult to infer gender-based preferences
simply by counting the number of same- and mixed-gender coauthorships.
Essentially, whenever coauthorship data are sampled from two or more
discrete sets of literature, which vary in the author gender ratio and
which are largely not connected by collaboration, the number of
same-gendered coauthors will be inflated. This can give the impression
that authors preferentially publish with same-gendered colleagues even
if no gender preferences exist, or if the true preference is for
opposite-gendered colleagues (`gender heterophily'). For example, a
sample of literature containing bioinformatics and cell biology papers
will probably contain an excess of mostly-male and mostly-female author
lists, simply because researchers usually collaborate within their own
discipline, and because the author gender ratio is more male-biased in
bioinformatics than in cell biology {[}5{]}.

\begin{figure}[htbp]
\centering
\includegraphics{../figures/Fig1.pdf}
\caption{The Wahlund effect can make it appear as if authors prefer to
publish with same-gendered colleagues, even if no such preference
exists. Here, coloured circles represent male and female authors, and
coauthors are linked with lines. Across the whole set of ten papers,
there is an apparent excess of same-gender collaborations: there are six
same-gender papers and only four mixed-gender papers, which is fewer
than the \(10\times2\times0.5\times0.5 = 5\) mixed-gender papers
expected under the null hypothesis that authors assort randomly with
respect to gender. However, within each subset, there is no evidence
that authors prefer to publish with same-gendered individuals (if
anything, this small dataset suggests gender heterophily). The Wahlund
effect will tend to inflate the frequency of same-sex coauthorships
whenever the data is composed of two or more disconnected subsets of
literature with different author gender ratios; these subsets could be
research disciplines, older versus newer papers, or papers from authors
in different countries. The example countries and disciplines were
selected based on {[}5{]}. \label{wahlund_plot}}
\end{figure}

In the present study, we test whether life sciences researchers tend to
co-publish with same-gendered colleagues, while controlling for the
Wahlund effect as strictly as possible. We use a recently-published
dataset describing the gender of 35.5m authors from 9.15m articles
indexed on PubMed {[}5{]}. Holman et al. {[}5{]} reported large
differences in the gender ratio of authors across research disciplines,
journals, countries, and across the years 2002-2016. We therefore tested
for gender homophily while restricting our analysis to particular
journals (i.e.~research specialties), time periods, and countries. We
quantified gender assortment using a metric called \(\alpha'\) {[}60{]},
which is positive when same-gender authors publish together more often
than expected (gender homohily), negative when opposite-gender authors
publish together more often than expected (heterophily), and equal to
zero when authors assort randomly with respect to gender (see Methods).

\section{Results}\label{results}

\subsection{Gender homophily by discipline, time period, and authorship
position}\label{gender-homophily-by-discipline-time-period-and-authorship-position}

\autoref{alpha_histograms} shows the distribution of \(\alpha'\)
estimates in 2015-2016 across all journals for which we recovered
sufficient data, when \(\alpha'\) was calculated for all authors, first
authors only, or last authors only. Most journals had positive values of
\(\alpha'\) (77-92\%, depending on time period and author type; S1
Data), and for many of these the FDR-corrected p-values suggested that
\(\alpha'\) was significantly greater than zero (1469/2077 journals were
significant in 2015-16, and 404/1192 in 2005-6; S1 Data). Only 2/2077
journals had statistically significantly heterophily (i.e.
\(\alpha' < 0\)) in 2015-16, and 1/1192 in 2005-6 (S2 Table). The
remaining 606 or 787 journals (in 2015 and 2005 respectively) had a
value of \(\alpha'\) not significantly different from zero, consistent
with the null hypothesis of random assortment with respect to gender. We
also confirmed that the majority of papers had multiple authors, in most
journals (S2 Data) and disciplines (S3 Data, S1 Fig).

\begin{figure}[htbp]
\centering
\includegraphics{../figures/Fig2.pdf}
\caption{Of the 2116 journals for which we had adequate data in
2015-2016, 825 showed statistically significant evidence of gender
homophily (denoted by \(\alpha' > 0\)), and 1 showed statistically
significant evidence of heterophily (\(\alpha' < 0\)), after false
discovery rate correction. The white area shows the number of journals
for which homophily was significantly stronger than expected under the
null hypothesis (corrected p \textless{} 0.05), while the blue area
shows all the remainder. Patterns were similar whether \(\alpha'\) was
calculated for all authors, for first authors only, or for last authors
only. \label{alpha_histograms}}
\end{figure}

@@

\subsection{Variance in homophily between
disciplines}\label{variance-in-homophily-between-disciplines}

\autoref{alpha_histograms} illustrates the variance in journal homophily
values (\(\alpha'\)) across scientific disciplines. All disciplines had
positive mean \(\alpha'\), although homophily appeared somewhat stronger
in some disciplines than others (e.g.~mean \(\alpha'\) was
\(0.12{\pm}0.02\) for Urology journals and \(0.03{\pm}0.01\) for
Veterinary Medicine journals; \autoref{alpha_histograms}, S4 Data).
However, there was no formal evidence for consistent differences in
\(\alpha'\) between disciplines: the random factor `Discipline'
explained around 1\% of the variance in \(\alpha'\) in the two linear
mixed models described in the previous section (see
\autoref{alpha_histograms} and mixed models in Online Supplementary
Material). Thus, the processes responsible for producing positive
\(\alpha'\) values appear to be similarly strong in all the disciplines
we examined.

There was no indication that journals publishing on a wide range of
topics have higher \(\alpha'\) values than more specialised journals,
due to the Wahlund effect. For example, the journal category
`Multidisciplinary' -- which includes journals like \emph{PLoS ONE},
\emph{Nature}, \emph{Science}, and \emph{PNAS} -- did not have notably
elevated \(\alpha'\) (\autoref{alpha_histograms}). This result suggests
that our estimates of homophily, and estimates from some earlier
studies, are not notably inflated by the presence of disparate research
topics (with variable author gender ratios) being published within
individual journals.

\subsection{Relationship between gender homophily and number of
authors}\label{relationship-between-gender-homophily-and-number-of-authors}

Papers with two authors had significantly lower (but still positive)
\(\alpha'\) values relative to papers with more than two authors, on
average (\autoref{author_number}; statistical results in Online
Supplmentary Material). Papers with 3, 4 or 5+ authors had essentially
identical average \(\alpha'\) values. The variance in \(\alpha'\) across
journals was also a little higher for 2-authors papers compared to the
remainder (\autoref{author_number}), though part of this variance is due
to the reduced sample size (in terms of number of authors) for the
2-author papers.

\begin{figure}[htbp]
\centering
\includegraphics{../figures/Fig3.pdf}
\caption{The coefficient of homophily (\(\alpha'\)) was slightly less
positive when calculated for two-author papers only, relative to papers
with longer author lists. The individual points, whose distribution is
summarised by the violin plots, correspond to individual journals. The
larger white points show the mean for each group (and its 95\% CIs), as
calculated by a Bayesian meta-regression model accounting for repeated
measures of \(\alpha'\) within journals, as well as the precision with
which \(\alpha'\) was estimated. \label{author_number}}
\end{figure}

\subsection{Relationship between gender homophily and gender
ratio}\label{relationship-between-gender-homophily-and-gender-ratio}

We next tested whether researchers are more or less likely to publish
with same-gendered colleagues in strongly gender-biased disciplines
(e.g.~Surgery or Nursing), relative to disciplines with a comparatively
gender-balanced workforce (e.g.~Psychiatry). We found a positive,
non-linear relationship between the overall gender ratio of all authors
publishing in a particular journal {[}5{]}, and the estimated value of
\(\alpha'\) for all authors and for first authors
(\autoref{alpha_gender_ratio}). Journals with a balanced or
female-biased author gender ratio tended to have higher \(\alpha'\) than
journals with a male-biased author gender ratio (GAM smooth terms p
\textless{} 0.001; Online Supplementary Material). The relationship was
not statistically significant when \(\alpha'\) was calculated for last
authors (GAM, p = 0.142), though the trend appeared similar
(\autoref{alpha_gender_ratio}).

\begin{figure}[htbp]
\centering
\includegraphics{../figures/Fig4.pdf}
\caption{There is a weakly positive, non-linear relationship between the
gender ratio of authors publishing in a journal, and the coefficient of
homophily (\(\alpha'\)). Specifically, journals with 50\% women authors
or higher tended to have more same-sex coauthorships than did journals
with predominantly men authors. This relationship held whether
\(\alpha'\) was calculated for all authors, first authors only, or last
authors only. A negative value on the x-axis denotes an excess of men
authors, a positive value denotes an excess of women authors, and zero
denotes gender parity. The lines were fitted using generalised additive
models with the smoothing parameter \(k\) set to 3.
\label{alpha_gender_ratio}}
\end{figure}

\subsection{Relationship between journal impact factor and gender
homophily}\label{relationship-between-journal-impact-factor-and-gender-homophily}

@@

\begin{figure}[htbp]
\centering
\includegraphics{../figures/Fig5.pdf}
\caption{Journal impact factor (expressed relative to the average for
the discipline) is negatively correlated with \(\alpha'\). The
relationship is noisy (\(R^2\) = 0.043), but the results suggest that
journals with strong homophily tend to have lower impact factors than
journals with weak homophily in the same discipline.
\label{impact_factor}}
\end{figure}

\subsection{Analysis accounting for differences in author gender ratio
between
countries}\label{analysis-accounting-for-differences-in-author-gender-ratio-between-countries}

@@

\subsection{\texorpdfstring{Theoretical expectations for \(\alpha\) when
the gender ratio differs between career
stages}{Theoretical expectations for \textbackslash{}alpha when the gender ratio differs between career stages}}\label{theoretical-expectations-for-alpha-when-the-gender-ratio-differs-between-career-stages}

As shown in \autoref{simulation}, we predict that \(\alpha\) is expected
to be non-zero, even if collaborators are randomly selected with respect
to gender, provided that there is a gender gap between career stages.
The extent to which \(\alpha\) deviates from zero depends on the
relative frequencies of collaboration within and between career stages.
When \textgreater{}50\% of collaborations were between early and
established researchers, we expect gender heterophily (\(\alpha < 0\)).
Conversely, when \textgreater{}50\% of collaborations occured within
career stages, we expect gender homophily (\(\alpha > 0\)). In a few
parameter spaces (shown in red; \autoref{simulation}), \(\alpha\) was
quite high, and overlapped with the values that we estimated
(\autoref{alpha_histograms}).

\begin{figure}[htbp]
\centering
\includegraphics{../figures/Fig6_inkscape.pdf}
\caption{When there is a difference in gender ratio between early-career
and established researchers, and collaboration is non-random with
respect to career stage, the null expectation for \(\alpha\) deviates
from zero. An excess of collaborations between career stages gives the
appearance of gender heterophily (lower rows, blue areas), while an
excess of within-career stage collaborations produced apparent gender
homophily (upper rows, red areas). However, the conditions required for
strong gender homophily are quite restrictive, making it unlikely that
this issue explains all of the homophily observed in Figure 2. Contour
lines denote increments of 0.1. \label{simulation}}
\end{figure}

Despite this overlap, \autoref{simulation} suggests that our main
conclusions (and those of other studies of gender homophily) are
probably robust to this career stage issue. We only expect strongly
positive \(\alpha\) when A) the gender ratio is highly skewed across
career stages (e.g.~a 5-fold difference), and B) collaborations between
early and established researchers are very rare (e.g. \textless{}10\% of
the total). Both of these conditions are untrue for most fields: the
gender gap across careers stages is generally less pronounced {[}1,5{]},
and it is very common for early-career researchers to co-publish with an
established mentor {[}61{]}. However, one can get \(\alpha > 0\) for
realistic combinations of parameters, e.g.~a moderate shortage of women
in senior positions coupled with a moderate excess of within-career
stage collaboration, suggesting this effect might contribute to some of
the observed homophily (in this and previous studies).

Lastly, we note that if there is a gender gap between career stages and
coauthorships between early-career and established researchers comprise
\textgreater{}50\% of the total, then the baseline expectation for
\(\alpha\) is actually less than zero (blue areas in
\autoref{simulation}). Therefore, our results might under-estimate the
extent to which researchers preferentially select same-gendered
collaborators in some cases.

\section{Discussion}\label{discussion}

>>>>

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Researchers collaborate with same-gendered colleagues more often than expected across the life sciences},
            pdfauthor={Luke Holman* and Claire Morandin; *;},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\setlength{\droptitle}{-2em}

  \title{Researchers collaborate with same-gendered colleagues more often than
expected across the life sciences}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{Luke Holman* and Claire Morandin\textsuperscript{§} \\ *\textit{luke.holman@unimelb.edu.au} \\ \textsuperscript{§}\textit{claire.morandin@helsinki.fi} \vspace{5mm}}
    \preauthor{\centering\large\emph}
  \postauthor{\par}
    \date{}
    \predate{}\postdate{}

@@

\begin{document}
\maketitle
\begin{abstract}
Evidence suggests that women in academia are hindered by conscious and
unconscious biases, and often feel excluded from formal and informal
opportunities for research collaboration. In addition to ensuring
fairness and helping to redress gender imbalance in the academic
workforce, increasing women's access to collaboration could help
scientific progress by drawing on more of the available human capital.
Here, we test whether researchers tend to collaborate with same-gendered
colleagues, using more stringent methods and a larger dataset than in
past work. Our results reaffirm that researchers co-publish with
colleagues of the same gender more often than expected by chance, and
show that this `gender homophily' is slightly stronger today than it was
10 years ago. Contrary to our expectations, we found no evidence that
homophily is driven mostly by senior academics, and no evidence that
homophily is stronger in fields where women are in the minority.
Interestingly, journals with a high impact factor for their discipline
tended to have comparatively low homophily, as predicted if mixed-gender
teams produce better research. We discuss some potential causes of
gender homophily in science. \vspace{5mm}

\par

\noindent \textbf{Keywords:} Gender bias, Homophily, Research
collaboration, Text mining, Women in STEM.
\end{abstract}

\maketitle\unmarkedfntext{*School of BioSciences, The University of Melbourne, Victoria, Australia.}
\maketitle\unmarkedfntext{\textsuperscript{§}Organismal and Evolutionary Biology Research Programme,
Faculty of Biological and Environmental Sciences, University of Helsinki, Finland.}

@@

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

@@

However, we believe that earlier studies of gender homophily were
hindered by a largely unacknowledged statistical issue that we will
refer to as the Wahlund effect (\autoref{wahlund_plot}), by analogy with
the conceptually similar Whalund effect in population genetics {[}59{]}.
The Wahlund effect makes it deceptively difficult to infer gender-based
preferences simply by counting the number of same- and mixed-gender
coauthorships. Essentially, whenever coauthorship data are sampled from
two or more discrete sets of literature, which vary in the author gender
ratio and which are largely not connected by collaboration, the number
of same-gendered coauthors will be inflated. This can give the
impression that authors preferentially publish with same-gendered
colleagues even if no gender preferences exist, or if the true
preference is for opposite-gendered colleagues (`gender heterophily').
For example, a sample of literature containing a mixture of
bioinformatics and cell biology papers will probably contain an excess
of mostly-male and mostly-female author lists, simply because
researchers usually collaborate within their own discipline, and because
the author gender ratio is more male-biased in bioinformatics than in
cell biology {[}5{]}.

\begin{figure}
\centering
\includegraphics{../figures/Fig1.pdf}
\caption{The Wahlund effect can make it appear as if authors publish
with same-gendered colleagues disproportionately often, even if
collaboration is completely random with respect to gender. Here,
coloured circles represent male and female authors, and coauthors are
linked with lines. Across the whole set of ten papers, there is an
apparent excess of same-gender collaborations: there are six same-gender
papers and only four mixed-gender papers, which is fewer than the
\(10\times2\times0.5\times0.5 = 5\) mixed-gender papers expected under
the null hypothesis that authors assort randomly. However, within each
subset, there is no evidence that authors prefer to publish with
same-gendered individuals (if anything, this small dataset suggests
gender heterophily). The Wahlund effect will tend to inflate the
frequency of same-gender coauthorships whenever the data is composed of
two or more disconnected subsets of literature with different author
gender ratios; these subsets could be research disciplines, older versus
newer papers, or papers from authors in different countries. The example
countries and disciplines were selected based on data in {[}5{]}.
\label{wahlund_plot}}
\end{figure}

In the present study, we test whether life sciences researchers tend to
co-publish with same-gendered colleagues, while controlling for the
Wahlund effect as strictly as possible. We use a recently-published
dataset describing the gender of 35.5 million authors from 9.15 million
articles indexed on PubMed {[}5{]}. Holman et al. {[}5{]} reported large
differences in the gender ratio of authors across research disciplines,
journals, countries, and across the years 2002-2016. We therefore tested
for gender homophily while restricting our analysis to particular
journals (i.e.~research specialties), time periods, and countries. We
quantified gender assortment using a metric called \(\alpha'\) {[}60{]},
which is positive when same-gender authors publish together more often
than expected (gender homophily), negative when opposite-gender authors
publish together more often than expected (heterophily), and equal to
zero when authors assort randomly with respect to gender (see Methods).

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{gender-homophily-by-discipline-time-period-and-authorship-position}{%
\subsection{Gender homophily by discipline, time period, and authorship
position}\label{gender-homophily-by-discipline-time-period-and-authorship-position}}

\autoref{alpha_histograms} shows the distribution of \(\alpha'\)
estimates in 2015-2016 across all journals for which we recovered
sufficient data, when \(\alpha'\) was calculated for all authors, first
authors only, or last authors only. Most journals had positive values of
\(\alpha'\) (77-92\%, depending on time period and author type; S1
Data), and for many of these the false discovery rate (FDR)-corrected
p-values suggested that \(\alpha'\) was significantly greater than zero
(1469/2077 journals were significant in 2015-16, and 404/1192 in 2005-6;
S1 Data). Only 2/2077 journals had statistically significant heterophily
(i.e. \(\alpha' < 0\)) in 2015-16, and 1/1192 in 2005-6 (S2 Table). The
remaining 606 or 787 journals (in 2015 and 2005 respectively) had a
value of \(\alpha'\) not significantly different from zero, consistent
with the null hypothesis of random assortment with respect to gender. We
also confirmed that in most journals (S2 Data) and most research
disciplines (S3 Data, S1 Fig), the majority of papers had multiple
authors. To verify the argument put forward in Figure 1, we also
calculated \(\alpha\) across all two-author papers in our entire 15-year
PubMed dataset (excluding single-author papers, and papers where at
least one author's gender was unknown; n = \textgreater{}3 million
papers, \textgreater{}16 million authors). Across the whole dataset,
\(\alpha\) was 0.126, which is almost double the median value for
individual journals (Figure 2; \(\alpha = 0.070\) for `All authors'),
suggesting that lumping together papers from different fields and
different time periods can indeed inflate \(\alpha\).

\begin{figure}
\centering
\includegraphics{../figures/Fig2.pdf}
\caption{Of the 2116 journals for which we had adequate data in
2015-2016, 825 showed statistically significant evidence of gender
homophily (denoted by \(\alpha' > 0\)), and 1 showed statistically
significant evidence of heterophily (\(\alpha' < 0\)), after false
discovery rate correction. In the stacked desnity plot, the white area
shows the number of journals for which homophily was significantly
stronger than expected under the null hypothesis (corrected p
\textless{} 0.05), while the blue area shows all the remainder. Patterns
were similar whether \(\alpha'\) was calculated for all authors, for
first authors only, or for last authors only. Points in the right panel
show \(\alpha'\) for individual journals.\label{alpha_histograms}}
\end{figure}

@@

\hypertarget{variance-in-homophily-between-disciplines}{%
\subsection{Variance in homophily between
disciplines}\label{variance-in-homophily-between-disciplines}}

\autoref{alpha_histograms} illustrates the variance in journal homophily
values (\(\alpha'\)) across scientific disciplines. All disciplines had
positive mean \(\alpha'\) (averaged over journals), although homophily
appeared somewhat stronger in some disciplines than others (e.g.~mean
\(\alpha'\) was \(0.12{\pm}0.02\) for Urology journals and
\(0.03{\pm}0.01\) for Veterinary Medicine journals;
\autoref{alpha_histograms}, S4 Data). However, there was no formal
evidence for consistent differences in \(\alpha'\) between disciplines:
the random factor `Discipline' explained around 1\% of the variance in
\(\alpha'\) in the two linear mixed models described in the previous
section (see \autoref{alpha_histograms} and mixed models in Online
Supplementary Material). Thus, whatever processes are responsible for
producing positive \(\alpha'\) values appear to be similarly strong in
all the disciplines we examined.

There was no indication that journals publishing on a wide range of
topics have higher \(\alpha'\) values than more specialised journals,
due to the Wahlund effect. For example, the journal category
`Multidisciplinary' -- which includes journals like \emph{PLOS ONE},
\emph{Nature}, \emph{Science}, and \emph{PNAS} -- did not have markedly
elevated \(\alpha'\) (\autoref{alpha_histograms}). This result suggests
that our estimates of homophily, and estimates from some of the earlier
studies listed in the Introduction, are probably not inflated by the
presence of disparate research topics (with variable author gender
ratios) being published within individual journals.

\hypertarget{relationship-between-gender-homophily-and-number-of-authors}{%
\subsection{Relationship between gender homophily and number of
authors}\label{relationship-between-gender-homophily-and-number-of-authors}}

Papers with two authors had significantly lower (but still positive)
\(\alpha'\) values relative to papers with more than two authors, on
average (\autoref{author_number}; statistical results in Online
Supplmentary Material). Papers with 3, 4 or 5+ authors had essentially
identical average \(\alpha'\) values. The variance in \(\alpha'\) across
journals was also a little higher for 2-authors papers compared to the
remainder (\autoref{author_number}), though part of this variance is due
to the reduced sample size (in terms of number of authors) for the
2-author papers. One possible explanation for this finding is that
2-authors papers might be more likely to have an author list that is
evenly split between career stages (e.g.~a postgraduate student and
their supervisor), increasing the chance that the authors are mixed
gender (see Figure 6). The result also suggests that the processes
responsible for gender homophily are similar in small (e.g.~3 author)
and larger (5+ authors) collaborations (and across disciplines where
small versus large collaborations are the norm).

\begin{figure}
\centering
\includegraphics{../figures/Fig3.pdf}
\caption{The coefficient of homophily (\(\alpha'\)) was slightly less
positive when calculated for two-author papers only, relative to papers
with longer author lists. The individual points, whose distribution is
summarised by the violin plots, correspond to individual journals. The
larger white points show the mean for each group (and its 95\% CIs), as
calculated by a Bayesian meta-regression model accounting for repeated
measures of \(\alpha'\) within journals, as well as the precision with
which \(\alpha'\) was estimated. \label{author_number}}
\end{figure}

\hypertarget{relationship-between-gender-homophily-and-gender-ratio}{%
\subsection{Relationship between gender homophily and gender
ratio}\label{relationship-between-gender-homophily-and-gender-ratio}}

We next tested whether researchers are more or less likely to publish
with same-gendered colleagues in strongly gender-biased disciplines
(e.g.~Surgery or Nursing), relative to disciplines with a comparatively
gender-balanced workforce (e.g.~Psychiatry). We found a positive,
non-linear relationship between the overall gender ratio of all authors
publishing in a particular journal {[}5{]}, and the estimated value of
\(\alpha'\) for all authors and for first authors
(\autoref{alpha_gender_ratio}). Journals with a balanced or
female-biased author gender ratio tended to have higher \(\alpha'\)
(i.e.~stronger homophily) than journals with a male-biased author gender
ratio (GAM smooth terms p \textless{} 0.001; Online Supplementary
Material). The relationship was not statistically significant when
\(\alpha'\) was calculated for last authors (GAM, p = 0.142), though the
trend appeared similar (\autoref{alpha_gender_ratio}).

\begin{figure}
\centering
\includegraphics{../figures/Fig4.pdf}
\caption{There is a weakly positive, non-linear relationship between the
gender ratio of authors publishing in a journal, and the coefficient of
homophily (\(\alpha'\)). Specifically, journals with 50\% women authors
or higher tended to have more same-sex coauthorships than did journals
with predominantly men authors. This relationship held whether
\(\alpha'\) was calculated for all authors, first authors only, or last
authors only. A negative value on the x-axis denotes an excess of men
authors, a positive value denotes an excess of women authors, and zero
denotes gender parity (i.e.~equal numbers of men and women). The lines
were fitted using generalised additive models with the smoothing
parameter \(k\) set to 3. \label{alpha_gender_ratio}}
\end{figure}

\hypertarget{relationship-between-journal-impact-factor-and-gender-homophily}{%
\subsection{Relationship between journal impact factor and gender
homophily}\label{relationship-between-journal-impact-factor-and-gender-homophily}}

@@

\begin{figure}
\centering
\includegraphics{../figures/Fig5.pdf}
\caption{Journal impact factor (expressed relative to the average for
the discipline) is negatively correlated with \(\alpha'\). The
relationship is noisy (\(R^2\) = 0.043), but the results suggest that
journals with strong homophily tend to have lower impact factors than
journals with weak homophily in the same discipline.
\label{impact_factor}}
\end{figure}

\hypertarget{analysis-accounting-for-differences-in-author-gender-ratio-between-countries}{%
\subsection{Analysis accounting for differences in author gender ratio
between
countries}\label{analysis-accounting-for-differences-in-author-gender-ratio-between-countries}}

@@

\hypertarget{theoretical-expectations-for-alpha-when-the-gender-ratio-differs-between-career-stages}{%
\subsection{\texorpdfstring{Theoretical expectations for \(\alpha\) when
the gender ratio differs between career
stages}{Theoretical expectations for \textbackslash{}alpha when the gender ratio differs between career stages}}\label{theoretical-expectations-for-alpha-when-the-gender-ratio-differs-between-career-stages}}

Given that we cannot identify individual researchers or their career
stages, we used a simple model to derive the theoretical expectations
for \(\alpha\) when the gender ratio differs between career stages. As
shown in \autoref{simulation}, we predict that \(\alpha\) is expected to
be non-zero, even if collaborators are randomly selected with respect to
gender, provided that there is a gender gap between career stages. The
extent to which \(\alpha\) deviates from zero depends on the relative
frequencies of collaboration within and between career stages (rows and
columns in Figure 6), and the size of the gender gap between stages (x-
and y-axes in Figure 6). When \textgreater{}50\% of coauthor pairs
comprise one early-career and one established researcher, we expect
gender heterophily (\(\alpha < 0\)) whenever the gender ratio differs
between career stages. Conversely, when \textgreater{}50\% of
collaborations are between people at the same career stage, we expect
gender homophily (\(\alpha > 0\)). In a few parameter spaces (shown in
red; \autoref{simulation}), \(\alpha\) was quite high, and overlapped
with the values that we estimated (\autoref{alpha_histograms}).

\begin{figure}
\centering
\includegraphics{../figures/Fig6_inkscape.pdf}
\caption{When there is a difference in gender ratio between early-career
and established researchers, and collaboration is non-random with
respect to career stage, the null expectation for \(\alpha\) deviates
from zero. An excess of collaborations between career stages gives the
appearance of gender heterophily (lower rows, blue areas), while an
excess of within-career stage collaborations produced apparent gender
homophily (upper rows, red areas). However, the conditions required for
strong gender homophily are quite restrictive, making it unlikely that
this issue explains all of the homophily observed in Figure 2. Contour
lines denote increments of 0.1. \label{simulation}}
\end{figure}

Despite this overlap, \autoref{simulation} suggests that our main
conclusions (and those of other studies of gender homophily) are
probably robust to this career stage issue. We only expect strongly
positive \(\alpha\) when A) the gender ratio is highly skewed across
career stages (e.g.~a 5-fold difference), and B) collaborations between
early and established researchers are very rare (e.g. \textless{}10\% of
the total). Both of these conditions seem unlikely to be true for most
fields: the gender gap across careers stages is generally less
pronounced {[}1,5{]}, and it is very common for early-career researchers
to co-publish with an established mentor {[}61{]}. However, one can get
\(\alpha > 0\) for realistic combinations of parameters, e.g.~a moderate
shortage of women in senior positions coupled with a moderate excess of
within-career stage collaboration, suggesting this effect might
contribute to some of the homophily observed by this and previous
studies.

Lastly, we note that if there is a gender gap between career stages and
coauthorships between early-career and established researchers comprise
\textgreater{}50\% of the total, then the baseline expectation for
\(\alpha\) is actually less than zero (blue areas in
\autoref{simulation}). Therefore, it is possible that researchers
preferentially assort with same-gendered collaborators even more
strongly than implied by other results, at least for certain journals or
research disciplines.

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}
